---
output: pdf_document
---
--- 
title: "Power and sample size calculations for interaction analysis"
output:
  html_document:
    toc: true
    theme: united
    number_sections: true
    highlight: default
---
<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Power and sample size calculations for interaction analysis}
-->

In this package power and sample size calculations for the analysis of treatment-by-subgroup interactions are proposed. Therefore, the interaction effect is either defined via [product-type interaction contrasts][producttypeinteraction] [1,2] or as [ratio of treatment differences][ratiosquantitative] [3]. The latter formulation allows besides the detection of interactions also the assessment of the consistency/heterogeneity of the subgroup-specific treatment effect by the definition of an inconsistency margin. he methodology is applicable in trials with a [continuous][ratiosquantitative] as well as a [binary endpoint][ratiosbinary].

* calculate the power to detect an interaction effect 

* calculate sample size to detect an interaction effect

* calculate power for consistency assessment

* calculate sample size for consistency assessment

#Consistency assessment in multi-regional clinical trials
The function `PowCon` calculates the minimal power (any-pair power) associated with simultaneous tests for the ratios of treatment differences with a user defined inconsistency margin for the assessment of treatment-by-subgroup interactions.

```{r, cache=TRUE, echo=TRUE, eval=FALSE, tidy=TRUE}
#library(poco)
PowCon(mu=c(10,15,10,15,10,15,10,15,10,11), 
       #vector of cell means from the cell 
       #mu = means model corresponding to a two-way layout
       sd=sqrt(5),
       #number samples per treatment-by-subgroup combination
       n = 10,
       #pooled standard deviation
       n.sub=5,
       #number of levels of the subgrouping factor; 
       #if n.sub=1 the argument SubMat is ignored
       TreatMat= "Tukey", 
       #type of contrast for the treatment factor; 
       #naming one of the contrast types available in contrMat(multcomp)
       SubMat = "GrandMean",
       #type of contrast for the subgrouping factor; 
       #naming one of the contrast types available in contrMat(multcomp); 
       #if n.sub=1 the argument SubMat is ignored
       thetas=1, #inconsistency margin(s)
       alpha=0.05, #familywise type I error to be controlled
       alternative="two.sided",#specifying the direction of the alternative hypothesis, 
       #one of "two.sided","less","greater"
       type="anypair")#power definition, one of "global","anypair","allpair"
```

The function `nPowCon`calculates the sample size associated with simultaneous tests for the ratios of treatment differences with a user defined inconsistency margin for the assessment of treatment-by-subgroup interactions. The required size n is determined iteratively by starting with a given sample size and search until the power condition is satisfied.

```{r, cache=TRUE, echo=TRUE, eval=FALSE, tidy=FALSE}
nPowCon(min.power=0.9,#any pair power
       mu=c(10,15,10,15,10,15,10,15,10,11),#vector of cell means from the 
       #cell means model corresponding to a two-way layout
       sd=sqrt(5),#pooled standard deviation
       n.sub=5,#number of levels of the subgrouping factor; 
       #if n.sub=1 the argument SubMat is ignored
       TreatMat= "Tukey", #type of contrast for the treatment factor; 
       #naming one of the contrast types available in contrMat(multcomp)
       SubMat = "GrandMean",#type of contrast for the subgrouping factor; 
       #naming one of the contrast types available in contrMat(multcomp); 
       #if n.sub=1 the argument
       thetas=1, #inconsistency margin(s)
       alpha=0.05, #familywise type I error to be controlled
       alternative="two.sided",##specifying the direction of the alternative hypothesis, 
       #one of "two.sided","less","greater"
       type="anypair")#power definition, one of "global","anypair","allpair"
```

#Qualitative interaction
Testing for or against a qualitative interaction is relevant in randomized clinical trials that use a common primary factor treatment and have a secondary factor, such as the centre, region, subgroup, gender or biomarker. [Kitsche and Hothorn][ratiosquantitative] formulated interaction contrasts for ratios of differences between the levels of the primary treatment factor to detect a qualitative interaction. To calculate the power and sample size to detect a qualitative interaction the user has to set the inconsistency margin `thetas` to 1 and specify the `alternative` argument to `greater`. The approach is also applicable to detect treatment-by-subset interactions in a stratified, randomised clinical trial with a binary-response variable, see [Kitsche (2014)][ratiosbinary]. The function `PowConBinom` calculates the any-pair power for multiple ratios of treatment differences for a binomial distributed endpoint. As illustrative example consider the [Metoprolol Controlled-Release Randomized Intervention Trial in Heart Failure][MERIT-HF] [5]. In this trial significant qualitative interactions were of particular interest, especially significant departures from the overall effect among any of the participating countries. The following code calculates the power to detect a qualitative interaction in this trial.

```{r, cache=TRUE, echo=TRUE, eval=FALSE, tidy=FALSE}
data(MetoCRXL2)#the data set is available from the poco package
library(MCPAN)#required package to estimate the success probabilities
MCPAN_Est <- binomest(Success ~ RegionTreat,data=MetoCRXL2, success="1", method="Wald")
MCPAN_Est$estp#estimated success probabilities
MCPAN_Est$n#sample size per treatment-by-region 
PowConBinom(p=MCPAN_Est$estp,
            #binomial proportions for each tretament-by-subgroup combination 
            n=MCPAN_Est$n, #sample sizes per treatment-by-subgroup combination
            n.sub = 12, #number of levels of the subgrouping factor; 
            #if n.sub=1 the argument SubMat is ignored
            TreatMat = "Tukey", #type of contrast for the treatment factor; 
            #naming one of the contrast types available in contrMat(multcomp)
            SubMat = "GrandMean",#type of contrast for the subgrouping factor; 
            #naming one of the contrast types available in contrMat(multcomp); 
            #if n.sub=1 the argument SubMat is ignored
            rhs = 0, #inconsistency margin(s)
            alternative = "less", #direction of the alternative hypothesis, 
            #one of "two.sided","less","greater"
            alpha = 0.05,#familywise type I error to be controlled
            type="anypair")#power definition, one of "global","anypair","allpair"
```


#Non-inferiority analysis
The presented methodology to assess the consistency of treatment effects is in general an extension of the problem of multiple testing for non-inferiority based on ratios was first adressed by [Hauschke and Kieser][Hauschke]. If the number of levels of the subgrouping factor is set to 1 the method simplifies to the approach of [Dilba et al.][Dilba] [7]. The following code recalculates the method presented in [Dilba et al.][Dilba].

```{r, echo=TRUE, results='hide', cache=TRUE ,eval=FALSE, tidy=FALSE}
PowCon(mu=c(10,9,9,11), 
       sd=sqrt(5),
       n = 10,
       n.sub=1,
       TreatMat= "Dunnett", 
       SubMat = "Tukey",
       thetas=0.8, 
       alpha=0.05, 
       alternative="greater",
       type="anypair")
```


#References

1. Gabriel KR, Putter J,Wax Y. Simultaneous Confidence Intervals for product-type Interaction Contrasts. *Journal of the Royal Statistical Society Series B - Statistical Methodology* **1973**; 35(2):234-244.
2. Kitsche A, Schaarschmidt F. Analysis of Statistical Interactions in Factorial Experiments. *Journal of Agronomy and Crop Science* **2014**; early view
3. Kitsche A, Hothorn LA. Testing for qualitative interaction using ratios of treatment differences. *Statistics in Medicine* **2014**; 13(9):1477-1489
4. Kitsche A. Detecting qualitative interactions in clinical trials with binary responses. *Pharmaceutical statistics* **2014**; early view
5. MERIT-HF Study Group. Effect of metoprolol CR/XL in chronic heart failure: Metoprolol CR/XL Randomised Intervention Trial In-congestive Heart Failure (MERIT-HF). *Lancet* **1999**; 353(9169):2001-2007.
6. Kitsche A, Power and sample size computations for simultaneous consistency assessment of treatment effects [manuscript in preparation]
7. Dilba G., Bretz F., Hothorn L.A., Guiard V. Power and sample size computations in simultaneous tests for non-inferiority based on relative margins. *Statistics in Medicine* **2006**; 25: 1131-1147


[producttypeinteraction]: http://onlinelibrary.wiley.com/doi/10.1111/jac.12076/abstract
[ratiosquantitative]:http://onlinelibrary.wiley.com/doi/10.1002/sim.6048/abstract
[ratiosbinary]: http://onlinelibrary.wiley.com/doi/10.1002/pst.1632/abstract
[MERIT-HF]:http://www.sciencedirect.com/science/article/pii/S0140673699044402
[Dilba]:http://onlinelibrary.wiley.com/doi/10.1002/sim.2359/abstract
[Hauschke]:http://onlinelibrary.wiley.com/doi/10.1002/(SICI)1097-0258(19990115)18:1%3C93::AID-SIM992%3E3.0.CO;2-8/abstract